InnoDB Full text Search hacking with Shiny/R

### 2015-12-31 article.rmd
Pushed more details and contents

```{r init, message = FALSE, echo=FALSE}
#options(width = 9999)
#knitr::opts_chunk$set(width=9999)
# opts_chunk$set(fig.width=9999) #, fig.height=6))

@@

> You only need to have a MySQL sandbox with the 5.7 version.

@@

A couple of days ago one of our customers came up with a question regarding FTS over InnoDB engine. Althuogh the question
is not entirely related with most of the content written here, I came up with the conclusion that FTS sometimes is a misunderstood feature.

The point of this article will be clarify when each algorithm fits in each situation.

Both official documentation and articles are propagated over internet, however sometimes most of them lacks to answer the
main questions and test cases. In order to follow a reproducible article, I used public available data sources (www.gutenberg.org).

In order to show the effects off the field sizes on the ranking algorithms, you will see two main tables (bookContent and 
bookContentByLine) which contains the books parsed by paragraph and by line respectively. You'll see the noise generated
by the `QUERY EXPANSION` algorithm when phrases are too large, and that's why you see both tables.

The current article has been developed using Shiny/R in order to allow you to see the effects of the algorithms.

@@

@@

SELECT  content, group_concat(it.POSITION) as positions, 
        round(MATCH(content) AGAINST ("country" IN NATURAL LANGUAGE MODE WITH QUERY EXPANSION)) as QERank, 
        round(MATCH(content) AGAINST ("country*" IN BOOLEAN MODE)) as BooleanRank, 
        length(content) as len 
      FROM bookContentByLine bl join information_schema.INNODB_FT_INDEX_TABLE it 
          ON (bl.FTS_DOC_ID = it.DOC_ID)  
      WHERE  MATCH(content) AGAINST ("country" IN NATURAL LANGUAGE MODE WITH QUERY EXPANSION) 
          AND lower(it.WORD) like '%country%' group by FTS_DOC_ID ORDER BY 3 DESC LIMIT 10 ;

+---------------------------------------------------------------------------+-----------+--------+-------------+------+
| content                                                                   | positions | QERank | BooleanRank | len  |
+---------------------------------------------------------------------------+-----------+--------+-------------+------+
| "country may be ennobled, and under its auspices may be verified that"    | 1         |     54 |           4 |   70 |
| "country, under a most excellent president, wherein all cities had their" | 1         |     53 |           4 |   73 |
| "Again, the prince who holds a country differing in the above respects"   | 31        |     51 |           4 |   71 |
| "country districts. As Lucca had five gates, he divided his own country"  | 1,63      |     51 |           8 |   72 |
| "inflicting immense damage upon the country. Whereupon the Florentines"   | 36        |     50 |           4 |   71 |
| "The cities of Germany are absolutely free, they own but little country"  | 64        |     49 |           4 |   72 |
| "citizens of Fermo, to whom the slavery of their country was dearer than" | 49        |     48 |           4 |   73 |
| "is useful in two ways. Firstly, he learns to know his country, and"      | 55        |     46 |           4 |   68 |
| "agree that he should retain any lordship over the country. Because the"  | 51        |     45 |           4 |   72 |
| "his own parts of the country, according to the authority he had assumed" | 22        |     43 |           4 |   73 |
+---------------------------------------------------------------------------+-----------+--------+-------------+------+

@@

```
| "country districts. As Lucca had five gates, he divided his own country"  | 1,63      |     51 |           8 |   72 |
```

@@

(test) > SELECT FTS_DOC_ID, 
                group_concat(it.POSITION) as positions, 
                round(MATCH(content) AGAINST ("country" IN NATURAL LANGUAGE MODE WITH QUERY EXPANSION)) as QERank, 
                round(MATCH(content) AGAINST ("country*" IN BOOLEAN MODE)) as BooleanRank, 
                length(content) as len 
          FROM bookContent bl join information_schema.INNODB_FT_INDEX_TABLE it ON (bl.FTS_DOC_ID = it.DOC_ID)  
          WHERE  MATCH(content) AGAINST ("country" IN NATURAL LANGUAGE MODE WITH QUERY EXPANSION) 
              AND lower(it.WORD) like '%country%' group by FTS_DOC_ID ORDER BY QERank DESC LIMIT 10 ;
+------------+------------------+--------+-------------+------+
| FTS_DOC_ID | positions        | QERank | BooleanRank | len  |
+------------+------------------+--------+-------------+------+
|        757 | 6858             |   2804 |           1 | 8101 | <<< This one on the top does only 1 exact match.
|        735 | 2842             |    925 |           1 | 3133 |
|        157 | 747              |    698 |           1 | 2385 |
|        733 | 265,1342,816,750 |    569 |           2 | 1765 |
|        739 | 1171             |    540 |           1 | 1659 |
|        249 | 197,739          |    435 |           1 | 1368 |
|        279 | 664              |    428 |           1 | 1365 |
|        411 | 468,410          |    396 |           2 | 1170 |
|        155 | 31,495,158,525   |    363 |           5 | 1378 |  <<< This doc contains 5 exact matches!
|        327 | 60               |    362 |           1 | 1372 |
+------------+------------------+--------+-------------+------+
10 rows in set (4,09 sec)
```

This kind of searches is useful if you want to find entries with words that appear frequently within your search.

@@

Following stats are from the bookContent table (which has an average length of `r rs$len` characters ).

```
(test) > select * 
          from information_schema.INNODB_FT_INDEX_TABLE 
          WHERE DOC_ID = 733 AND lower(WORD) like '%country%';
+------------------+--------------+-------------+-----------+--------+----------+
| WORD             | FIRST_DOC_ID | LAST_DOC_ID | DOC_COUNT | DOC_ID | POSITION |
+------------------+--------------+-------------+-----------+--------+----------+
| country          |          149 |         787 |        28 |    733 |      265 |
| country          |          149 |         787 |        28 |    733 |     1342 |
| countrydistricts |          733 |         733 |         1 |    733 |      816 |
| thecountry       |          249 |         733 |         2 |    733 |      750 |
+------------------+--------------+-------------+-----------+--------+----------+
4 rows in set (0,08 sec)

@@

- Use QUERY EXPANSION only if you are interested to search relations over exact matches. Remember that the field
  size is crucial when using this.
- FTS is not the best fit for exact string matches. You won't use FTS for searching emails or strings as words using
  single fields. 
- If you are using BOOLEAN MODE, you can use the rank score to filter rows. MySQL is clever enough to optimize the 
  FTS functions to avoid double executions. You can do this using somehting like:
  `match(content,title) against ("first (<second >third)") > 1 `
  Generally, scores lower than 1 can be ignored when using boolean or natural mode searches. 
- `OPTIMIZE TABLE` does a rebuild of the table. To avoid this, set `innodb_optimize_fulltext_only=1` in order to do an incremental
  maintance on the table.
- Recall that NATURAL LANGUAGE MODE does not take the operands as the BOOLEAN MODE.
- If you plan to order by rank, it is not necessary to specify the clause `ORDER BY` as MySQL does the order forcelly internally. Also,
  the behavior is different from the default as it returns the heaviest at the top (like ORDER BY rank DESC).
- If you come from MyISAM's FTS implementation, recall that the ranking scoring is different.
- Create the FULLTEXT index after the data is loaded [InnoDB bulk load](http://dev.mysql.com/doc/refman/5.7/en/optimizing-innodb-bulk-data-loading.html). When restoring FTS backups, you will probably hit the "ERROR 182 (HY000) at line nn: Invalid InnoDB FTS Doc ID".
- Try to don't use more than one FTS expression in the where clause. Keep in mind that this affects the order in the results and
  it consumes a considerably amount of CPU. MySQL orders by the latest expression in the WHERE clause. [WL#7123](https://dev.mysql.com/worklog/task/?id=7123)
- Also, if avoiding the rank information in the projection (SELECT clause) and using other aggregations like count(*), will use
  the "no ranking" FT_hints. The `limit` hint won't be used if invoked explicitely an `ORDER BY` and th `MATCH` clause in the projection.

>>>>

```{r init, message = FALSE, echo=FALSE}
library(png)
#options(width = 9999)
#knitr::opts_chunk$set(fig.width=9999)
# opts_chunk$set(fig.width=9999) #, fig.height=6))

@@

If you want to run this locally, please install a MySQL 5.7 instance and:

- Uncomment the line __source('~/fts_article/load.R')__ and run the report, or
- Restore the dumps located in the `dumps` folder. Uncomment the table creation.

@@

A couple of days ago one of our customers came up with a question regarding FTS
over InnoDB engine. Although the question is not answered in the current article,
I came up with the conclusion that FTS is sometimes misunderstood.

The point of this "article" is to show dynamically how the search algorithms work,
using non-fictional data (data sources were downloaded from www.gutenberg.org
project) within an easy interface.

In order to show the effects off the field sizes over the query expansion algorithm,
you will see two main tables (bookContent and bookContentByLine) both containing
the same books in different approaches: by line and by paragraph. You'll see the
noise generated by the `QUERY EXPANSION` algorithm when phrases are too large.

The current article has been developed using Shiny/R in order to allow you to see
the effects of the algorithms. It connects to a open database at `db4free.net`,
which unfortunately is not 100% available.

## Action!

@@

To start showing the effects of the algorithms, the following example searches
the word 'country' using `query expansion`. This means that we are not looking
only the exact matches, but also the entries that appear the most when the
the exact match has been found.

In the `SELECT` clause you'll see both FTS expressions using `NATURAL LANGUAGE`
with query expansion and `BOOLEAN`modes respectively.

@@

SELECT  content, bookid, group_concat(it.POSITION) as pos,
        round(MATCH(content) AGAINST ("country" IN NATURAL LANGUAGE MODE WITH QUERY EXPANSION)) as QERank,
        round(MATCH(content) AGAINST ("country" IN BOOLEAN MODE)) as BoolRank
      FROM bookContentByLine bl join information_schema.INNODB_FT_INDEX_TABLE it
          ON (bl.FTS_DOC_ID = it.DOC_ID)  
      WHERE  MATCH(content) AGAINST ("country" IN NATURAL LANGUAGE MODE WITH QUERY EXPANSION)
          AND it.WORD = 'country'
     GROUP BY FTS_DOC_ID
     ORDER BY 4 DESC
     LIMIT 10 ;

+-------------------------------------------------------------------------------+--------+------+--------+----------+
| content                                                                       | bookid | pos  | QERank | BoolRank |
+-------------------------------------------------------------------------------+--------+------+--------+----------+
| "country in September, migrating into Holland, and leave their mates behind"  |  15707 | 1    |    105 |        7 |
| "unsatisfied desire to serve his country, the two prevalent enthusiasms at"   |  15707 | 33   |     98 |        7 |
| "Language, Vol. I. p. 212. In this country, where four or five horses travel" |  15707 | 35   |     93 |        7 |
| "inflicting immense damage upon the country. Whereupon the Florentines"       |   1232 | 36   |     89 |        7 |
| "made for a country of twenty or thirty millions’ population, can be laid"  |  39064 | 12   |     89 |        7 |
| "The spiders of this country manufacture nets of various forms, adapted to"   |  15707 | 21   |     87 |        7 |
| "a velvet-covered arm-chair at my head! This country is too decadent"         |  33415 | 45   |     86 |        7 |
| "country may be ennobled, and under its auspices may be verified that"        |   1232 | 1    |     84 |        7 |
| "name. The writer of this unpublished pamphlet sees his country in a"         |  39064 | 56   |     84 |        7 |
| "In our country, Mr. Pennant informs us, that some quails migrate, and"       |  15707 | 8    |     83 |        7 |
| "all the morning in passing over the adjacent country." (Voyage to Senegal,"  |  15707 | 46   |     82 |        7 |
| "the electoral system of the country. Immediately an outcry burst out"        |  39064 | 29   |     82 |        7 |
| "country, under a most excellent president, wherein all cities had their"     |   1232 | 1    |     81 |        7 |
| "Though in this country horses shew little vestiges of policy, yet in the"    |  15707 | 16   |     81 |        7 |
| "country districts. As Lucca had five gates, he divided his own country"      |   1232 | 1,63 |     80 |       14 |
+-------------------------------------------------------------------------------+--------+------+--------+----------+
15 rows in set (1,16 sec)

@@

```
| "country districts. As Lucca had five gates, he divided his own country" |   1232 | 1,63 |     80 |       14 |
```

@@

(test) > SELECT bookid, FTS_DOC_ID,
                group_concat(it.POSITION) as positions,
                round(MATCH(content) AGAINST ("country" IN NATURAL LANGUAGE MODE WITH QUERY EXPANSION)) as QERank,
                round(MATCH(content) AGAINST ("country" IN BOOLEAN MODE)) as BooleanRank,
                length(content) as len
          FROM bookContent bl join information_schema.INNODB_FT_INDEX_TABLE it ON (bl.FTS_DOC_ID = it.DOC_ID)  
          WHERE  MATCH(content) AGAINST ("country" IN NATURAL LANGUAGE MODE WITH QUERY EXPANSION)
              AND it.WORD = 'country'
          GROUP BY FTS_DOC_ID
          ORDER BY QERank DESC
          LIMIT 10 ;

+--------+------------+-----------------+--------+-------------+-------+
| bookid | FTS_DOC_ID | positions       | QERank | BooleanRank | len   |
+--------+------------+-----------------+--------+-------------+-------+
|  16452 |      17637 | 942,2552,9084   |  32494 |          10 | 51790 |
|  16452 |      17827 | 31699           |  30232 |           3 | 51701 |
|  16452 |      17761 | 667,47646       |  29517 |           7 | 50264 |
|  16452 |      17791 | 13566           |  28888 |           3 | 49129 |
|  16452 |      17927 | 23259,7044      |  26731 |           7 | 48983 |
|  16452 |      17839 | 9012,199        |  24933 |           7 | 44451 |
|  16452 |      17815 | 29318           |  24745 |           3 | 44011 |
|  16452 |      17729 | 895,16485,24034 |  23305 |          10 | 42612 |
|  16452 |      17621 | 1765            |  19935 |           3 | 36698 |
|  16452 |      17803 | 3942            |  17552 |           3 | 30586 |
+--------+------------+-----------------+--------+-------------+-------+
10 rows in set (1,88 sec)
```

@@

The query expansion is useful when you intend to search which entries contain
more words that appear frequently within the search term. Having large text fields
increase the probability to have more words that appear among the search term.
In the case of `bookContent` table (by paragraph table), the average field size
is `r rs$len` characters.

## The INNODB_FT_INDEX_TABLE

There is a way to play with the contents of the FTS indexes. As you ma noticed
in the previos examples, I used the `set global innodb_ft_aux_table = 'test/bookContent';`
statement, which loads the index content to memory for an easy querying.

i.e. You can easily get the most frequent tokens:

```
(test) > select WORD,count(*)
           FROM information_schema.INNODB_FT_INDEX_TABLE   
           group by WORD having count(*) > 1000
           order by 2
           limit 10;
+--------+----------+
| WORD   | count(*) |
+--------+----------+
| should |     1023 |
| yet    |     1027 |
| any    |     1070 |
| like   |     1071 |
| been   |     1073 |
| first  |     1080 |
| nor    |     1087 |
| your   |     1106 |
| thou   |     1130 |
| shall  |     1164 |
+--------+----------+
10 rows in set (5,40 sec)
```

Probably it isn't a very useful information as most of this words appears too frequently
and are modal verbs,  adverbs,  pronouns, determiners, etc.

It could be the case that you are not interested on indexing those words. If that's
the case you can add them as `stopwords` in your own stopwords table. I built I
query for this situation to allow us to build the stopwords table using the current
words that we want to add to the filtering:

```
(test) > select group_concat(WORD) FROM (select distinct WORD  
  FROM information_schema.INNODB_FT_INDEX_TABLE               
  group by WORD having count(*) > 1000) d\G
*************************** 1. row ***************************
group_concat(WORD): all,and,any,been,but,can,first,had,has,have,her,him,his,into,
its,like,may,more,nor,not,now,one,only,other,our,said,shall,she,should,some,such,
than,their,them,then,there,these,they,those,thou,thus,thy,time,were,which,would,
yet,you,your
1 row in set (5,28 sec)

Let's build our filter table using both default and new entries and keeping the
alphabetical order:

```
CREATE TABLE bookContentByLine_stopwords(value VARCHAR(30)) ENGINE = INNODB;

INSERT INTO bookContentByLine_stopwords
SELECT value FROM (
    SELECT value FROM
    INFORMATION_SCHEMA.INNODB_FT_DEFAULT_STOPWORD
    UNION
    SELECT DISTINCT WORD as value  
      FROM information_schema.INNODB_FT_INDEX_TABLE               
      GROUP BY WORD having count(*) > 1000
) allEntries
ORDER BY value ASC;

DROP INDEX ftscontent ON bookContentByLine;
SET GLOBAL innodb_ft_server_stopword_table = 'test/bookContentByLine_stopwords';
CREATE FULLTEXT INDEX ftscontent ON bookContentByLine(content);

TODO: Check index sizes

```
(test) > select *
          from information_schema.INNODB_FT_INDEX_TABLE
          WHERE WORD = 'country';
+------------------+--------------+-------------+-----------+--------+----------+
| WORD             | FIRST_DOC_ID | LAST_DOC_ID | DOC_COUNT | DOC_ID | POSITION |
+------------------+--------------+-------------+-----------+--------+----------+
| country          |          149 |         787 |        28 |    733 |      265 |
| country          |          149 |         787 |        28 |    733 |     1342 |
| countrydistricts |          733 |         733 |         1 |    733 |      816 |
| thecountry       |          249 |         733 |         2 |    733 |      750 |
+------------------+--------------+-------------+-----------+--------+----------+
4 rows in set (0,08 sec)

@@

- Use QUERY EXPANSION only if you are interested to search relations over exact matches. Remember that the field
  size is crucial when using this.
- FTS is not the best fit for exact string matches. You won't use FTS for searching emails or strings as words using
  single fields.
- If you are using BOOLEAN MODE, you can use the rank score to filter rows. MySQL is clever enough to optimize the
  FTS functions to avoid double executions. You can do this using somehting like:
  `match(content,title) against ("first (<second >third)") > 1 `
  Generally, scores lower than 1 can be ignored when using boolean or natural mode searches.
- `OPTIMIZE TABLE` does a rebuild of the table. To avoid this, set `innodb_optimize_fulltext_only=1` in order to do an incremental
  maintance on the table.
- Recall that NATURAL LANGUAGE MODE does not take the operands as the BOOLEAN MODE.
- If you plan to order by rank, it is not necessary to specify the clause `ORDER BY` as MySQL does the order forcelly internally. Also,
  the behavior is different from the default as it returns the heaviest at the top (like ORDER BY rank DESC).
- If you come from MyISAM's FTS implementation, recall that the ranking scoring is different.
- Create the FULLTEXT index after the data is loaded [InnoDB bulk load](http://dev.mysql.com/doc/refman/5.7/en/optimizing-innodb-bulk-data-loading.html). When restoring FTS backups, you will probably hit the "ERROR 182 (HY000) at line nn: Invalid InnoDB FTS Doc ID".
- Try to don't use more than one FTS expression in the where clause. Keep in mind that this affects the order in the results and
  it consumes a considerably amount of CPU. MySQL orders by the latest expression in the WHERE clause. [WL#7123](https://dev.mysql.com/worklog/task/?id=7123)
- Also, if avoiding the rank information in the projection (SELECT clause) and using other aggregations like `count(*)`, will use
  the "no ranking" FT_hints. The `limit` hint won't be used if invoked explicitely an `ORDER BY` and th `MATCH` clause in the projection.
